Uncovering the Heterogeneity and Ontogeny of Mouse Thymic Macrophages Reveals an Unexpected Early Checkpoint Role

Version 1

Abstract

Thymic macrophages (TMs) are critical for maintaining thymus homeostasis, yet their heterogeneity and specific functions in T cell development remain unclear. Through single-cell RNA sequencing and transgenic reporter mice, we revealed two novel TM subsets characterized by distinct surface markers and transcriptional programs. TIMD4⁺ VCAM1⁺ CX3CR1⁻ TMs were enriched for phagocytic and apoptotic-cell–clearance pathways, whereas TIMD4⁻ VCAM1⁺ CX3CR1⁺ TMs exhibited elevated expression of antigen-presentation machinery and interferon-response genes. Pseudotime trajectory analysis indicated a potential lineage progression from TIMD4⁻ VCAM1⁺ CX3CR1⁺ precursors toward TIMD4⁺ VCAM1⁺ CX3CR1⁻ effectors. Functional interrogation using fetal thymic organ cultures and the MaFIA depletion mouse model demonstrated that acute loss of TMs severely impaired thymocyte development, particularly at the pre-T cell receptor (TCR) β-selection checkpoint. TM ablation led to marked reductions in post-β-selection thymocyte subsets, implicating TMs in both survival and differentiation signals required for early T cell development. Mechanistically, TM depletion disrupted pre-TCR signaling and hindered transition through β-selection, likely due to absence of both apoptotic-cell clearance and differentiation cues provided by specialized TM subsets. These findings define two distinct TM populations and establish their role in orchestrating thymocyte maturation. By elucidating TM subset specialization and their dynamic contributions to the β-selection process, our findings advance our understanding of thymic microenvironment interactions and highlights macrophage heterogeneity as a key regulator of T cell development.

Introduction

The thymus is the primary lymphoid organ dedicated to the generation of a diverse and self-tolerant repertoire of T lymphocytes, a process fundamental to adaptive immunity (1). This intricate T cell developmental program unfolds within a highly organized microenvironment composed of thymic epithelial cells (TECs), endothelial cells, and various hematopoietic populations, which collectively provide the essential signals for thymocyte differentiation, proliferation, and selection (2–4). The developmental journey of a T cell begins when bone marrow-derived thymic seeding progenitors (TSPs) enter the thymus at the cortico-medullary junction (CMJ) and commit to the T cell lineage, initiating a multi-stage differentiation cascade (5–7).

This journey is punctuated by a series of stringent checkpoints that assess the integrity and functionality of the nascent T cell receptor (TCR) repertoire (8,9). The earliest stages occur in the thymic cortex, where progenitors, lacking expression of CD4 and CD8, termed double-negative (DN) thymocytes (6), start to differentiate, which is characterized by the expression of CD44 and CD25 (3,4). Progression from the DN1 (CD44+CD25-) to the DN2 (CD44+CD25+) and DN3 (CD44-CD25+) stages is marked by the initiation of V(D)J recombination at the Tcrb, Tcrg, and Tcrd gene loci (1). The DN3 stage represents the first critical checkpoint in αβ T cell development, known as β-selection (12,13). Here, thymocytes that have successfully rearranged a functional TCRβ chain pair it with the invariant pre-TCRα chain (pTα) to form the pre-TCR complex. Successful signaling through this complex provides vital survival, proliferation, and differentiation cues, allowing the cell to progress to the DN4 stage and subsequently to the CD4+CD8+ double-positive (DP) stage (14,15). Thymocytes failing to generate a functional pre-TCR are eliminated via apoptosis, ensuring that only cells with a viable TCRβ chain continue their developmental progression (14).

Following β-selection, DP thymocytes undergo further TCRα chain rearrangement and are subjected to two additional selection events. Positive selection ensures that the fully assembled αβ TCR can recognize self-peptides presented by major histocompatibility complex (MHC) molecules on cortical (c)TECs, a process essential for TCR MHC-restriction (16–18). Subsequently, negative selection in the medulla eliminates thymocytes whose TCRs bind with high affinity to self-peptides presented by medullary (m)TECs and dendritic cells (DCs), thereby establishing central tolerance (19,20). These selection processes are notably stringent, resulting in the apoptotic death of over 95% of all developing thymocytes (21).

Integral to this journey and processes are thymic macrophages (TMs), which have long been recognized for their critical homeostatic function. Residing in both the cortex and medulla, TMs are exceptionally efficient phagocytes, responsible for the swift and silent clearance of the millions of apoptotic thymocytes generated daily (22–25). This process, known as efferocytosis, is vital for preventing the release of potentially immunogenic and inflammatory intracellular contents and maintaining the integrity of the thymic architecture (26). However, this view of TMs as mere "housekeepers" is rapidly evolving. Recent advances, particularly single-cell RNA sequencing (scRNA-seq), have revealed significant heterogeneity within the TM compartment, suggesting functional specialization (27). Two principal subsets have been identified in the adult mouse thymus: cortical TIMD4+ macrophages, which have potent efferocytosis function, and CX3CR1+ macrophages, which are enriched at the CMJ and express genes associated with antigen presentation, suggesting at a potential role in negative selection (27).

Despite these advances, our understanding of the full spectrum of TM function remains incomplete. While a role in clearing the cellular debris from positive and negative selection is well-established, and a contribution to negative selection itself is hypothesized, the involvement of TMs in the earliest stages of T cell development has been largely overlooked. The β-selection checkpoint, a nexus of proliferation, survival, and lineage commitment, occurs in a microenvironment rich in TMs. Yet, whether these macrophages are passive bystanders or active participants in this critical decision-making process is unknown.

The objective of this study is to investigate the role of thymic macrophages in early T cell development, with a specific focus on the β-selection checkpoint. By integrating high-resolution transcriptomic analysis, fate-mapping models, and functional assays, we sought to dissect the interactions between distinct TM subsets and developing DN thymocytes. Our findings reveal a previously unappreciated function for thymic macrophages in actively shaping the outcome of β-selection, thereby influencing the survival and differentiation of early T cell precursors. This work redefines the role of thymic macrophages, expanding their function beyond efferocytosis to include direct participation in the foundational checkpoints of T cell repertoire formation.

Materials and methods

Animals

Macrophage Fas-Induced Apoptosis (MaFIA), MafB-mCherry-Cre, Spic−/−, and C57BL/6J mice were purchased from the Jackson Laboratory. The Flt3-Cre × Rosa-mTmG mice and Flt3-Cre × Rosa-mTmG mice in the Ccr2-/- background were kindly provided by Dr. Slava Epelman (University Health Network, University of Toronto) (28). All animals were bred and maintained in specific pathogen-free conditions at the Comparative Research Facility of the Sunnybrook Research Institute (SRI). All animal procedures were approved by the Sunnybrook Health Sciences Centre Animal Care Committee (Toronto, Ontario, Canada).

Timed pregnancies and embryonic thymus analysis

Timed pregnancies were set up by housing male mice (8–24 weeks of age) in separate cages and mating them with female mice (8–24 weeks of age) overnight. The following morning, females were separated from the males, and successful matings were identified by the presence of a vaginal plug. The gestational age of the embryos was designated as embryonic day 0.5 (E0.5) on the day the plug was observed. Thymuses were harvested from embryos at E15 and E17, as well as from postnatal mice at 1 day, 1 week, and 3 weeks of age, in addition to adult (6-8 weeks) mice. Following collection, thymuses were subjected to enzymatic digestion to dissociate the tissue into single-cell suspensions. The resulting cells were stained for flow cytometric analysis.

Intravenous CD45 staining

A CD45-Phycoerythrin (PE) antibody (30-F11; BioLegend) was diluted in phosphate buffered saline (PBS) to a final concentration of 1.5 µg/150 µL. The antibody solution was administered via tail vein injection using a 1 mL insulin syringe. Three minutes after the injection, mice were euthanized, and thymuses were harvested. The collected tissue was then processed and stained for flow cytometric analysis.

Cell isolation from thymus

Thymocytes were harvested and subjected to enzymatic digestion using the Spleen Dissociation Kit (Miltenyi Biotec) and the gentleMACS™ Dissociator (Miltenyi Biotec) according to the manufacturer’s instructions. The resulting cell suspensions were filtered through a 70 µm cell strainer to remove debris, resuspended in PBS, and kept on ice until further processing.

Flow cytometry

Single-cell suspensions (1 × 107 cells) were prepared from the thymus and stained in a 96-well conical bottom plate (ThermoFisher Scientific). The staining cocktail included the fixable viability dye eFluor 450 (eBioscience), diluted 1:800 in PBS, anti-CD16/32 to block Fc receptors, and the desired fluorochrome-conjugated antibodies targeting macrophages and T cells (see Supplemental Tables 1 and 2 for details). Cells were initially stained with eFluor 450 in PBS for 15 minutes on ice. After washing with fluorescence-activated cell sorting (FACS) buffer, anti-CD64, anti-CCR2, and anti-VCAM1 antibodies were added to the wells, and cells were stained for an additional 15 minutes on ice. To block non-specific binding, Fc-block antibody was added to the cells 15 minutes prior to the addition of the remaining antibody cocktail, which was incubated with the cells for 30 minutes on ice.

Following staining, cells were washed and resuspended in FACS buffer for flow cytometric acquisition using the FACSymphony™ A5 Cell Analyzer (BD Biosciences). Data were analyzed with FlowJo software (TreeStar). The complete list of antibodies and their corresponding clones is provided in Supplemental Tables 1 and 2.

Cell sorting

Thymuses were harvested from 6-week-old C57BL/6 mice and subjected to enzymatic digestion using the Spleen Dissociation Kit (Miltenyi Biotec) and the gentleMACS™ Dissociator (Miltenyi Biotec) to generate single-cell suspensions. After cell isolation, cells were counted and stained with allophycocyanin (APC)-conjugated anti-mouse CD64 antibody in FACS buffer for 20 minutes on ice (see Supplemental Table 1 for antibody details). Following a wash with FACS buffer, cells were stained with anti-APC MicroBeads (Miltenyi Biotec) for 20 minutes on ice. The cells were then washed again with FACS buffer and loaded onto LS columns (Miltenyi Biotec) for positive selection of CD64+ cells. After selection, the cells were washed and further stained with antibodies against F4/80, TIMD4, Ly6C, and CD45 for 30 minutes on ice (see Supplemental Table 1 for antibody details). The stained cells were sorted using the FACSAria™ Fusion Flow Cytometer (BD Biosciences). CD64+ sorted cells were collected in 1.5 mL Eppendorf tubes containing 300 µL of FBS, counted, and resuspended in PBS supplemented with 0.04% BSA. Cells were then encapsulated, and library preparation was performed using the Chromium Single Cell 3’ Reagent Kit (10X Genomics). Gene expression libraries were sequenced on a NovaSeq 6000 platform (Illumina) at Azenta Life Sciences. In total, 3,214 and 2,815 cells were sequenced from two independent experiments.

Fetal thymus organ culture (FTOC) and TM depletion in MaFIA mice

Timed pregnancies were set up for MaFIA mice as described above. Fetal thymuses were harvested at E15.5 for FTOC. Briefly, culture wells were set up in a 12-well-plate 24 hour before FTOC. In each well, Whatman® Nuclepore™ Track-Etched membrane (WHA110409; Sigma) was placed on top of SURGIFORM® absorbable gelatin sponge (1974; Ethicon) in 1.5 mL of complete Dulbecco’s Modified Eagle Medium (DMEM) (supplemented with 10% Fetal Bovine Serum, 2 mM L-glutamine, 100 U/mL penicillin and 100 μg/mL streptomycin) (29,30). Three fetal thymuses were placed on top of the membrane for each well. B/B homodimerizer stock was diluted to 2.5 μM in complete DMEM and added to the thymus lobes at the time of culture setup and again on the following day. On day 2, the membrane with the thymus lobes was transferred to a fresh culture plate containing new gelatin sponge and complete DMEM. Cultures were maintained in 37 °C incubator until day 6, at which point the thymic lobes were processed for flow cytometry analysis as described above.

Transcriptomic analysis

CD64+ cells were isolated from the thymus of C57BL/6J adult mice and processed for scRNA-seq as described below. Data from each scRNA-seq experiment underwent pre-processing as detailed below. Quality control filters were applied to remove cells with more than 12.5% mitochondrial gene expression, fewer than 500 reads, or more than 6,000 reads, ensuring the retention of high-quality cells for analysis. Macrophages and monocytes were identified using a module scoring approach, and their respective cell barcodes were recorded. These barcodes were used for subsetting before merging the datasets from the two scRNA-seq experiments with published TM data from Dzhagalov’s group. The merging was performed using Harmony to align the datasets and correct for batch effects. Following integration, cells were clustered, and module scores were assigned to all identified clusters. Non-relevant cell types, including B cells, T cells, and DCs, were excluded from subsequent analyses to focus on macrophages and monocytes.

Analysis of scRNA-Seq and data integration

All analyses were performed using R software version 4.4.1. Cells expressing fewer than 500 genes or more than 6,000 genes were excluded from the dataset. Additionally, cells with greater than 12.5% mitochondrial gene content were removed to ensure data quality. The dimensionality of the dataset was assessed using an Elbow plot to determine the appropriate number of principal components. Gene ontology (GO) enrichment analysis of differentially expressed genes (DEGs) was conducted to identify overrepresented biological processes using the FindMarkers function from the Seurat package (31). The data were then integrated with previously published scRNA-seq datasets on TMs (GEO: GSE185460), as well as datasets from heart, liver, and lung macrophages (GEO: GSE188647), from the laboratories of Drs. Dzhagalov and Epelman, respectively. Batch correction across datasets was performed using the Harmony package (27,32,33). Following integration, the merged dataset was normalized using the SCTransform method to account for technical variation. For data visualization, Uniform Manifold Approximation and Projection (UMAP) was applied to reduce dimensionality and enable visualization of cellular clustering patterns.

Trajectory analysis using Monocle3

Trajectory analysis was performed using Monocle3 to infer the developmental trajectory of TMs (34). Prior to this, TM datasets were normalized and integrated using the SCTransform method implemented in Seurat. From the resulting Seurat object, the expression matrix, cell metadata, and gene metadata were extracted to construct a CellDataSet (CDS) object, which serves as the foundational data structure in Monocle3. Pre-processing steps were applied to the CDS to ensure high-quality trajectory inference. Principal component analysis (PCA) was first conducted to reduce dimensionality while preserving variance across the dataset. This was followed by non-linear dimensionality reduction using Uniform Manifold Approximation and Projection (UMAP) to visualize the global structure of the dataset in a low-dimensional space. The cells were then re-clustered within Monocle3 to refine group assignments, ensuring that clusters corresponded accurately to biological populations. Trajectory inference was performed with the root node manually assigned to monocytes (defined by Ly6c2 and Ccr2 expression), reflecting their position as the presumptive progenitor population in the trajectory.

Statistical analysis

Data are presented as the mean ± standard error of the mean (SEM). Univariate comparisons between two groups were performed using the Mann-Whitney U test. For analyses involving more than two groups, the normality of the data was assessed using the D’Agostino-Pearson test and/or inspection of Q-Q plots, while homoscedasticity (equal variance) was evaluated using the Brown-Forsythe test or by examining residual plots. Non-parametric comparisons across multiple groups were conducted using the Kruskal-Wallis test, followed by post-hoc pairwise comparisons of the mean ranks of each group with the mean rank of the control group. Multiple comparisons were adjusted using Dunn’s test to control for type I error. For pairwise comparisons involving more than two groups, the Holm-Šídák method was applied to correct for multiple comparisons, reducing the likelihood of type I error. All statistical analyses were performed using GraphPad Prism version 10.

Results

Identification of Two Distinct Tissue-Resident Macrophage Subsets in the Adult Thymus

To investigate the heterogeneity of phagocytic cells within the thymus, we performed multi-parameter flow cytometry on hematopoietic cells (CD45+) from the thymus of adult mice. We first identified the total TM population based on the co-expression of the canonical markers CD64 and F4/80 (Figure 1A-B). The majority of these CD64+ F4/80+ cells also expressed the phagocytic receptor MerTK, consistent with their identity as macrophages (35).

TMs characterization using flow cytometry and immunofluorescent staining.

Frozen thymic sections were used for immunofluorescent staining to identify the localization of TMs. Thymuses were harvested from C57BL/6 mice aged six to ten months and enzymatically digested for single-cell preparation. Cells were stained with antibodies for flow cytometric analysis. A) Gating strategy for identifying TMs. B) Representative flow cytometry histogram showing MerTK expression on TMs (blue) compared to CD64-F4/80- lymphocytes (red). C) Identification of distinct TM subsets using TIMD4 and VCAM1 markers. D) Comparison of the percentages of TM subsets in six-month-old mice. E) Comparison of the percentages of MerTK+ cells between TM subsets. F) Representative flow cytometry plots showing TM reporter expressions in MafB-mCherry and Csf1r-EGFP mice. Data are presented as the mean ± SEM (n = 7 per group from 2 independent experiments). A Kruskal-Wallis test was performed for multiple comparisons between three groups (*p ≤ 0.05).

To resolve heterogeneity within this population, we further analyzed for the expression of TIMD4 and VCAM1. This revealed three distinct subsets within the CD64+ F4/80+ gate: a major TIMD4+ VCAM1+ population, a smaller TIMD4- VCAM1+ population, and a minor TIMD4-VCAM1- population (Figure 1C-E). To clarify the identity of these subsets, we utilized MafB-mCherry and Csf1r-EGFP reporter mice. Both the TIMD4+ VCAM1+ and TIMD4- VCAM1+ subsets expressed high levels of MafB (mCherry+), a transcription factor crucial for mature tissue-resident macrophages (TRMs), while the TIMD4- VCAM1- subset was MafB-negative and expressed high levels of CSF1R (GFP+), a marker associated with monocytes (Figure 1F).

To definitively distinguish these thymus resident populations from circulating cells and further define their identity, we performed intravenous (IV) labeling with an anti-CD45 antibody. Both the TIMD4+ VCAM1+ and TIMD4- VCAM1+ macrophage populations were largely protected from IV labeling (CD45-), confirming their parenchymal, tissue-resident localization (Figure 2A, B, C). Furthermore, these two subsets lacked expression of classical monocyte-associated markers, including Ly6C, CCR2, and CX3CR1 (Figure 2D, E). In contrast, the TIMD4- VCAM1- subset expressed high levels of these monocyte markers, confirming its identity as a population of thymic monocytes rather than mature macrophages.

Distinguishing TMs from thymic and circulating monocytes.

Thymuses were harvested from six-month-old C57BL/6 mice and enzymatically digested for single-cell preparation. Cells were stained with antibodies for flow cytometric analysis. A) CD45 IV labeling of circulating leukocytes in the mouse thymus, distinguishing intravascular cells from thymic-resident populations. B) Comparison of the percentages of CD45 labeled cells between Ly6C+CD11b+ and CD64+F4/80+ cells in the thymus. C) Comparison of the percentages of CD45 labeled cells between TM subsets. D) Lack of expression of monocyte markers associated with TIMD4+VCAM1+ and TIMD4-VCAM1+ subsets. E) Comparative analysis of monocyte marker expression across TM subsets. Data are presented as mean ± SEM (n = 5-7 per group from 2 independent experiments). Statistical significance was assessed using a two-tailed Mann-Whitney test (**p ≤ 0.01) for comparisons between two groups. A Kruskal-Wallis test was performed for multiple comparisons between three groups (*p ≤ 0.05 and **p ≤ 0.01).

Taken together, our analysis robustly identifies two distinct subsets of bona fide tissue-resident thymic macrophages, a dominant TIMD4+ VCAM1+ population and a TIMD4- VCAM1+ population, alongside a separate population of thymic monocytes. This phenotypic framework allows for the specific investigation of the functional roles of each macrophage subset in thymic biology.

Transcriptomic Profiling by scRNA-seq Confirms Two Functionally Distinct Macrophage Subsets

To gain deeper insight into the transcriptional identities of these macrophage populations, we performed scRNA-seq on sorted thymic CD45+ CD64+ cells. Unbiased clustering of the resulting 3,153 cells revealed five distinct myeloid clusters, including monocytes, proliferating macrophages, and two major quiescent macrophage populations that corresponded directly to our flow cytometry findings (Figure 3A).

Single-cell transcriptomics reveals distinct subpopulations and gene expression profiles of thymic macrophages.

Thymuses were harvested from six-week-old C57BL/6 mice, and CD64⁺ cells were isolated by FACS for scRNA-seq using the 10x Genomics 3’ RNA-seq platform. For TM scRNA-seq, cells were pooled from six mice per sample, with each experiment performed independently in duplicate. The resulting dataset was integrated with published data from Dzhagalov’s lab where they sequenced from Cd11cYFP and MaFIA mice using Harmony for batch correction and alignment. A) UMAP clustering of TMs. B) Heatmap representation of DEGs across the five clusters. C) Comparison of macrophage numbers in the thymus and spleen of SpiC-/- mice. Data are presented as mean ± SEM (n = 4-5 per group from 2 independent experiments), analyzed using a two-tailed Mann-Whitney test (*p ≤ 0.05).

A major macrophage cluster was defined by high expression of Timd4, Vcam1, Mertk, and the transcription factor Maf, confirming its identity as the TIMD4+ VCAM1+ population (Figure 3B). In contrast, the other major macrophage cluster was defined by high expression of Vcam1 and Cx3cr1, aligning it with the TIMD4- VCAM1+ subset (Table 1), as identified above. Intriguingly, the TIMD4+ VCAM1+ CX3CR1- cluster also showed uniquely high expression of Spic, the lineage-defining transcription factor for splenic red pulp macrophages (RPMs) (36). To test for a shared developmental dependency, we analyzed SpiC knockout (Spic−/−) mice. While Spic−/− mice exhibited the expected loss of RPMs, the numbers of both thymic macrophage subsets were unaltered, suggesting that TMs utilize a Spic-independent developmental pathway despite their transcriptional similarities (Figure 3C).

To formally assess the functional specialization suggested by these transcriptional profiles, we performed Gene Ontology (GO) pathway analysis. This revealed a stark functional dichotomy between the two subsets. The TIMD4- VCAM1+ CX3CR1+ TMs were significantly enriched for pathways related to "antigen processing and presentation via MHC class I" and "interferon signaling" consistent with their high expression of H2-K1, H2-D1, and Runx3 and suggesting a role in T cell education (Figure 4A). Conversely, the TIMD4+ VCAM1+ CX3CR1- TM cluster was dominated by pathways associated with "phagocytosis" and "endocytosis," supporting a primary role in efferocytosis for apoptotic cell clearance (Figure 4B).

GO pathway analysis of TMs.

Thymuses were harvested from six-week-old C57BL/6 mice, and CD64⁺ cells were isolated by FACS for scRNA-seq using the 10x Genomics 3’ RNA-seq platform. For TM scRNA-seq, cells were pooled from six mice per sample, with each experiment performed independently in duplicate. The resulting dataset was integrated with published data from Dzhagalov’s laboratory where they sequenced from Cd11cYFP and MaFIA mice using Harmony for batch correction and alignment. A) Dotplot showing top 20 enriched GO pathways in TIMD4-VCAM1+CX3CR1+ TM subset based on highly expressed genes. B) Dotplot showing top 20 enriched GO pathways in TIMD4+VCAM1+CX3CR1- TM subset based on highly expressed genes.

This functional role was confirmed by the identification of a sub-cluster within the TIMD4+ VCAM1+ CX3CR1- population that was actively engaged in efferocytosis in vivo. These cells co-expressed canonical T-cell transcripts such as Cd3g and Trac alongside their macrophage-specific genes (Figure 3B). Thus, their T-cell transcriptional signature provides direct evidence of ongoing efferocytosis, capturing macrophages in the act of engulfing apoptotic thymocytes and confirming the potent phagocytic activity of TIMD4+ VCAM1+ CX3CR1- TMs.

The Thymic Macrophage Compartment Undergoes Dynamic Remodeling During Development

To understand how the TM compartment evolves in parallel with the thymus itself, we analyzed the number and composition of TM subsets during ontogeny, from embryonic development through adulthood. Consistent with known thymic biology, the absolute number of total thymocytes and TMs peaked in young, 3-week-old mice, before declining with age, indicating that the size of the TM pool is coupled to the overall cellularity of the organ (Supplemental Figure 1A-B).

Further analysis of TM subset composition revealed a profound and dynamic remodeling process across the lifespan. At embryonic day 15 (E15), the nascent thymus was populated almost exclusively by macrophages with a progenitor-like phenotype (TIMD4− VCAM1−), which co-expressed the monocyte-associated markers CD11b and CCR2 (Figure 5A). This suggests that the embryonic thymus is initially seeded by monocyte-derived precursors. Following birth, this landscape was dramatically reshaped. We observed a rapid postnatal expansion of the TIMD4+ VCAM1+ CX3CR1- subset, which became the major TM population in young mice, coinciding with the period of peak thymic size (Figure 5B). This indicates that the establishment of the mature, homeostatic TM pool is a key feature of postnatal thymic development. Interestingly, as mice aged and the thymus began to involute (6 weeks and older), the proportion of TIMD4+ VCAM1+ CX3CR1- macrophages steadily declined. This decline was accompanied by a progressive accumulation of the TIMD4- VCAM1+ CX3CR1+ TM subset, which became a more prominent feature of the aging thymus.

Developmental progression of TM subsets.

Thymuses were harvested from mice at various developmental time points and enzymatically digested to prepare single-cell suspensions. Cells were stained with antibodies for flow cytometry analysis. A) Expression of TM markers across different age groups. B) Comparison of the numbers and percentages of TM in mice at different ages. Data are presented as mean ± SEM (n = 4-12 per group from 2-4 independent experiments).

Together, these data reveal that the thymic macrophage compartment is not a static entity but is actively remodeled throughout life. Distinct TM subsets dominate at specific developmental windows, with an initial wave of progenitor-like cells being replaced by a homeostatic TIMD4+ VCAM1+ CX3CR1- population, which in turn gives way to a TIMD4- VCAM1+ CX3CR1+ population during the start of the thymic involution process.

Thymic Macrophage Subsets Arise from Distinct Developmental Origins

Having established the dynamic nature of the TM compartment, we next sought to define the developmental origins of the two major macrophage subsets. We first utilized Flt3-Cre × Rosa-mTmG fate-mapping mice, in which cells passing through a FLT3-expressing hematopoietic stem cell (HSC)-derived progenitor stage are permanently labeled with GFP, while cells from an FLT3-independent pathway (i.e., yolk sac-derived) remain Tomato+ (Figure 6A) (28). This fate-mapping analysis revealed a clear divergence in origin. The majority (>50%) of the homeostatic TIMD4+ VCAM1+ CX3CR1- macrophages were Tomato+, indicating they arise from an FLT3-independent, likely embryonic/fetal, precursor and persist into adulthood (Figure 6B-C). In stark contrast, the vast majority of TIMD4- VCAM1+ CX3CR1+ TMs and thymic monocytes were GFP+, demonstrating their origin from adult HSC-derived, FLT3-dependent progenitors.

Ontology of TM subsets.

Thymuses were harvested from 11–14-week-old Flt3-Cre × Rosa-mTmG mice and enzymatically digested to obtain single-cell suspensions. Cells were stained with antibodies for flow cytometric analysis. A) Schematic representation of the Flt3-Cre × Rosa-mTmG reporter mouse model. B) Representative flow cytometry plots showing mTomato (Tomato) and mGFP (GFP) reporter expression across TIMD4+VCAM1+, TIMD4-VCAM1+, and TIMD4- VCAM1- subsets. C) Comparison of reporter expression percentages among TM subsets. Data are presented as mean ± SEM (n = 7 per group from 2 independent experiments). Statistical significance was assessed using the Mann-Whitney test for pairwise comparisons, followed by the Holm-Šídá method to correct for multiple comparisons (***p ≤ 0.001).

Based on these findings, we hypothesized that the maintenance of the TIMD4- VCAM1+ CX3CR1+TM subset, but not the TIMD4+ VCAM1+ CX3CR1- subset, would depend on the continuous recruitment of CCR2-dependent monocytes from the bone marrow. To test this, we analyzed TM populations in Ccr2 knockout (Ccr2−/−) mice. While overall thymocyte numbers were unaffected, the composition of the TM compartment was significantly altered (Figure 7A). While there was no change in TIMD4+ VCAM1+ CX3CR1- macrophages number, the TIMD4- VCAM1+ CX3CR1+macrophage population and thymic monocytes was significantly decreased in Ccr2−/− mice (Figure 7B-D).

TIMD4-VCAM1+ TM subset is HSC derived.

Thymuses were harvested from 11–14-week-old Flt3-Cre × Rosa-mTmG mice (Control) and Flt3-Cre × Rosa-mTmG mice in the CCR2 knockout background (Ccr2-/-), followed by enzymatic digestion to prepare single-cell suspensions. Cells were stained with antibodies for flow cytometric analysis. A) Representative flow cytometry plots showing TM subsets from Ccr2-/- and control mice. B) Comparison of TIMD4+VCAM1+ TM subset percentage and absolute numbers between Ccr2-/- and control mice. C) Comparison of TIMD4-VCAM1+ TM subset percentage and absolute numbers between Ccr2-/- and control mice. D) Comparison of TIMD4-VCAM1- TM percentage and absolute numbers between Ccr2-/- and control mice. Data are presented as mean ± SEM (n = 6–7 per group from 2 independent experiments). Statistical significance was assessed using a two-tailed Mann-Whitney test (*p ≤ 0.05, **p ≤ 0.01, and ***p ≤ 0.001).

Together, these data establish a dual-origin model for the thymic macrophage network: a stable, self-maintaining population of TIMD4+ VCAM1+ CX3CR1- macrophages established from fetal, FLT3-independent precursors, which coexists with a dynamic, CCR2-dependent population of TIMD4- VCAM1+ CX3CR1+ macrophages that is continuously replenished by circulating monocytes.

Depletion of Csf1r-Expressing Myeloid Cells in Thymic Organ Culture Impairs T Cell Development at the β-Selection Checkpoint

To directly test the functional requirement for TMs during T cell development, we employed the MaFIA (Csf1r-EGFP/Fas-FKBP) mouse model, which permits the inducible depletion of Csf1r-expressing cells (Supplemental Figure 2A-B) (37). Because both TM subsets and thymic monocytes express the Csf1r-EGFP reporter, this system enables the targeted depletion of the thymic myeloid compartment (Supplemental Figure 2C-E). To circumvent the systemic effects of in vivo depletion, we utilized an ex vivo fetal thymic organ culture (FTOC) system (38). Thymic lobes from E15.5 MaFIA embryos were cultured with the dimerizer drug AP20187 to induce apoptosis in Csf1r+ cells (Figure 8A).

FTOC analysis of TM function during T cell development in MaFIA mice.

Thymuses from E15.5 MaFIA mice were used to establish FTOCs. Cultures were treated with 2.5 mM AP20187 to deplete Csf1r-expressing cells on the day of setup and again the following day. On day 2, FTOC films were transferred to fresh media plates, and samples were analyzed by flow cytometry on day 6 to assess T cell development kinetics. A) Schematic representation of the TMs depletion workflow in FTOC culture. B) Percentages and absolute numbers of EPCAM⁺ cells in FTOC samples on day 2, comparing control and AP20187-treated groups, analyzed by flow cytometry. C) Total thymocyte numbers on day 6, comparing control and AP20187-treated groups. D) Absolute numbers of DCs on day 6, comparing control and AP20187-treated groups. DCs were gated as CD45+CD11c+MHC-IIhi. Data are presented as mean ± SEM (n = 8–11 per group from 3 independent experiments). Statistical significance was assessed using a two-tailed Mann-Whitney test (*p ≤ 0.05 and ***p ≤ 0.001). E) Thymuses from embryonic day 15.5 (E15.5) MaFIA (Csf1r-EGFP) and congenic C57BL/6 (45.2) mice were used to setup FTOCs. Cultures were treated with 2.5 mM AP20187 to deplete Csf1r-expressing cells on the day of setup and again the following day. On day 2, FTOC films were transferred to fresh media plates, and samples were analyzed by flow cytometry on day 6 to assess T cell development kinetics. Comparison of TM percentages between control and AP20187-treated MaFIA mice. F) Comparison of TM absolute numbers in control and AP20187-treated MaFIA FTOCs, as well as in AP20187-treated C57BL/6 FTOCs. G) Comparison of TIMD4+VCAM1+CX3CR1- numbers between control and AP20187-treated MaFIA mice. H) Comparison of TIMD4-VCAM1+CX3CR1+ numbers between control and AP20187-treated MaFIA mice. Data are presented as the mean ± SEM (n = 8–11 per group from 3 independent experiments). Statistical significance was assessed using a two-tailed Mann-Whitney test (*p ≤ 0.05 and **p ≤ 0.01) for comparisons between two groups. A Kruskal-Wallis test was performed for multiple comparisons between three groups (*p ≤ 0.05).

To validate the specificity of our approach, we first examined the non-myeloid stromal compartment. Flow cytometry analysis confirmed that the number of TECs, identified as EpCAM+ CD45−, appeared unaffected by the treatment (Figure 8B). In contrast, depletion was highly effective against the intended myeloid targets. We observed a significant reduction in the total number of thymocytes and thymic DCs (Figure 8C-D). Crucially, the treatment also resulted in the profound loss of total TMs (Figure 8E). This effect was specific to the MaFIA system, as treatment of control C57BL/6 mice with the dimerizer did not cause TM depletion (Figure 8F). Analysis of the TM populations revealed that both the TIMD4+ VCAM1+ CX3CR1- and the TIMD4- VCAM1+ CX3CR1+ subsets were significantly depleted (Figure 8G-H).

Depletion of Csf1r+ cells led to an impairment in αβ T cell development (Figure 9A). AP20187-treated thymic lobes exhibited a significant reduction in the number of CD4+ CD8+ DP thymocytes (Figure 9B-C). While the development of γδ T cells was unaffected (Figure 9D). To identify the origin of this defect, we analyzed the upstream DN progenitor stages (Figure 10A). This revealed a developmental block, as treated FTOCs showed a significant accumulation of DN3 (CD44− CD25+) thymocytes and a corresponding decrease in the subsequent DN4 (CD44− CD25−) population (Figure 10B).

Impairment of T cell development in the absence of TMs.

Thymuses from embryonic day 15.5 (E15.5) MaFIA mice were used to establish FTOCs. Cultures were treated with 2.5 mM AP20187 to deplete Csf1r-expressing cells on the day of setup and again the following day. On day 2, FTOC films were transferred to fresh media plates, and samples were analyzed by flow cytometry on day 6 to assess T cell development kinetics. A) Representative flow cytometry plot showing T cell populations defined by CD4 and CD8 markers. B) Comparison of percentages and absolute numbers of developing T cells between control and AP20187-treated FTOC samples. C) Ratio of CD3⁺CD4⁺ single-positive (CD4SP) to CD3⁺CD8⁺ single-positive (CD8SP) thymocytes in control versus AP20187-treated FTOCs. Data are presented as the mean ± SEM (n = 8–11 per group from 3 independent experiments). D) Comparison of percentage and absolute numbers of γδ T cells between control and AP20187-treated FTOC samples. Data are presented as the mean ± SEM (n = 6–7 per group from 2 independent experiments). Statistical significance was assessed using the Mann-Whitney test, followed by the Holm-Šídák method to correct for multiple comparisons (*p ≤ 0.05, **p ≤ 0.01, and ***p ≤ 0.001).

TMs are critical for β-selection during T cell development.

Thymuses from embryonic day 15.5 (E15.5) MaFIA mice were used to establish FTOCs. Cultures were treated with 2.5 mM AP20187 to deplete Csf1r-expressing cells on the day of setup and again the following day. On day 2, FTOC films were transferred to fresh media plates, and samples were analyzed by flow cytometry on day 6 to assess T cell development kinetics. A) Representative flow cytometry plot showing DN cell populations defined by CD44 and CD25 markers B) Comparison of percentages and absolute numbers of DN cells between control and AP20187-treated FTOC samples. Data are presented as mean ± SEM (n = 8–11 per group from 3 independent experiments). C) Comparison of CD27+ DN3 cells percentage between control and AP20187-treated FTOC samples. D) Comparison of CD27+ DN4 cells percentage between control and AP20187-treated FTOC samples. Data are presented as mean ± SEM (n = 6–8 per group from 2 independent experiments). Statistical significance was assessed using the Mann-Whitney test for pairwise comparisons, followed by the Holm-Šídá method to correct for multiple comparisons (*p ≤ 0.05, **p ≤ 0.01, and ***p ≤ 0.001).

To pinpoint the impairment more precisely, we examined the expression of CD27, a surface marker upregulated upon successful pre-TCR signaling. Within the arrested DN3 population of treated lobes, the proportion of CD27+ cells was significantly reduced (Figure 10C-D). This points to a failure of thymocytes to receive or transduce the signals required to pass the β-selection checkpoint. Collectively, these data reveal that the Csf1r-expressing myeloid compartment, which includes thymic macrophages, appears to be important for the successful transition of thymocytes through β-selection, a critical early checkpoint in T cell development.

Discussion

The thymus provides a unique and specialized niche for T cell development, yet the contributions of its resident myeloid populations have been historically underappreciated. This study redefines the role of TMs, moving beyond their established function as mere "housekeepers" to reveal them as a heterogeneous and dynamic population that also is integral for the earliest checkpoints in T cell development. By combining high-resolution phenotyping, fate-mapping, and functional depletion, we demonstrate that the thymus contains two distinct macrophage subsets with separate origins and functions, and we uncover a critical role for the myeloid compartment in supporting thymocyte progression through β-selection.

Our work addresses the complexity of the TM landscape by integrating flow cytometry and single-cell transcriptomics. We identified a dominant TIMD4+ VCAM1+ CX3CR1- population transcriptionally programmed for efferocytosis, expressing high levels of Mertk and sharing a gene signature (Spic, MafB) with splenic red pulp macrophages (36). However, using Spic knockout mice, we showed that TMs, unlike their splenic counterparts, follow a Spic-independent developmental pathway, highlighting their unique adaptation to the thymic niche. In contrast, the TIMD4- VCAM1+ CX3CR1+ subset was enriched for genes involved in antigen presentation and interferon signaling, positioning it as a candidate for shaping the T cell repertoire, a role previously attributed almost exclusively to DCs (20).

Furthermore, we establish a dual-origin model for this network. The homeostatic TIMD4+ VCAM1+ CX3CR1- macrophages are likely long-lived, self-maintaining cells derived from embryonic, FLT3-independent precursors. Conversely, the TIMD4- VCAM1+ CX3CR1+ subset is continuously replenished from HSC-derived, Flt3-dependent monocytes, representing a more dynamic population likely involved in immune surveillance (39,40). This clarifies conflicting reports in the literature and aligns TM biology with the broader principles of tissue-resident macrophage ontogeny observed in other organs (23,27).

An impactful finding of this study is the appreciation of novel role for thymic myeloid cells in early T cell development. Using an ex vivo FTOC system to avoid the confounding stress of in vivo depletion, we demonstrated that the loss of Csf1r-expressing cells leads to a profound developmental block at the DN3 to DN4 transition. The accumulation of DN3 cells, coupled with a specific reduction in CD27 expression, pinpoints a failure at the β-selection checkpoint (41,42). While traditionally viewed as a cell-autonomous event, our data strongly suggest that pre-TCR signaling is not independent of the microenvironment (12). This support from myeloid cells could be multifaceted, involving the presentation of peptide-MHC to the pre-TCR (43–45) or other receptor-ligand interactions, the provision of essential survival factors (46–48), or the maintenance of a non-inflammatory niche through the clearance of apoptotic cells (49).

Although our depletion strategy also affected DCs, several lines of evidence implicate TMs as the primary mediators of this effect. The β-selection checkpoint occurs in the thymic cortex, a region densely populated by TMs, whereas DCs are largely restricted to the medulla and cortico-medullary junction (50). Nevertheless, future studies using more targeted depletion strategies will be crucial to definitively dissect the relative contributions of TMs and DCs. Additionally, the absence of TMs might lead to an accumulation of apoptotic cells, creating a potentially hostile environment for early thymocyte development. Macrophages, through their role in clearing apoptotic cells, produce anti-inflammatory cytokines like IL-10 and TGF-β, while suppressing pro-inflammatory cytokines such as IL-12 (51–53). Without these regulatory influences, TM depletion may result in localized thymic inflammation, further compromising thymocyte survival and development (54). Together, these findings highlight a supportive role of TMs in β-selection, not only by facilitating essential pre-TCR interactions but also by preserving a supportive microenvironment for thymocyte maturation.

In conclusion, this study elevates thymic macrophages from simple scavengers to critical architects of the initial TCRβ repertoire. By defining their heterogeneity, origins, and functional specialization, we reveal their integral role in thymic homeostasis and uncover their essential contribution to the successful navigation of the earliest checkpoints in T cell development. These findings not only advance our fundamental understanding of thymic biology but also open new avenues for exploring how modulation of this myeloid niche could influence immune tolerance and T cell-based therapies.

eLife Assessment

This study characterizes the heterogeneity and developmental origins of macrophages in the thymus and offers tantalizing evidence of their potential involvement in the first step of T cell selection. The macrophage characterisation is interesting, although the evidence for the specific involvement of macrophages in beta-selection is incomplete, as alternative explanations have not been ruled out. These results provide an important advance that further our understanding of thymus biology, especially in view of the contribution of heterogenous thymic macrophage subpopulations.

Reviewer #1 (Public review):

Summary:

The current manuscript characterizes in detail the macrophages in the thymus. The authors identify two distinct populations of thymic macrophages and describe their surface marker expression and transcriptional signatures. They also explore their ontology and kinetics of settling and persistence in the thymus and find that the TIMD4+ macrophages are derived from embryonic progenitors and self-maintain in the thymus, while the TIMD4- macrophages are derived from monocytes. Most importantly, the authors test the functional importance of thymic macrophages for T cell development using an in vitro depletion system, from which they conclude that macrophages are important for one of the earliest selection steps in T cell development - the beta selection.

Strengths:

The authors use state-of-the-art techniques, such as multiple genetically modified mice, multi-color flow cytometry, single-cell RNA sequencing, genetic fate mapping, and fetal thymic organ culture (FTOC) combined with depletion. Their work is in good agreement with prior published studies on the subject, such as Tacke et al. (PMID: 26091486) and Zhou et al. (PMID: 36449334). In addition to reproducing prior knowledge, the authors uncover novel and unexpected facets of thymic macrophage biology, such as their SpiC independence and the fact that TIMD4- thymic macrophages depend on CCR2 (Tacke et al. have shown that the overall thymic macrophage compartment is normal in CCR2-/- mice). Most surprisingly, the authors claim that thymic macrophages control an early checkpoint in T cell development, the beta selection. This has not been reported before, as beta selection is usually considered a cell-autonomous process in thymocytes that does not require input from other cells.

Weaknesses:

The thymic macrophage depletion experiments are not well controlled, and the authors' interpretation of the results is a stretch. First, the treatment depletes other cell types, most notably dendritic cells (DCs), which have well-known roles in thymic selection (though not specifically in beta selection). The authors' reasoning that macrophages are abundant in the cortex, where beta selection occurs, while DCs are enriched in the medulla, seems questionable, as the embryonic thymus typically lacks (or has very small) medulla. A second salient point is that the authors haven't ruled out direct toxicity of the dimerizer drug AP20187 on thymocytes (specifically DN cells) in MAFIA mice.

Altogether, this is a solid manuscript that largely confirms the previously established ontogeny and heterogeneity of thymic macrophages. However, the participation of thymic macrophages in beta selection needs stronger evidence.

Reviewer #2 (Public review):

This manuscript from Zuniga-Pflucker laboratory describes that thymic macrophages are heterogeneous in flow cytometric and transcriptomic profiles, containing two major populations characterized by TIMD4 and CX3CR1 expression. These macrophage populations are both parenchymal in the thymus but are unequal in developmental ontogeny, Flt3 expression history, and CCR2 dependency. The manuscript further reports the interesting findings that the depletion of thymic macrophages impairs thymocyte development at the DN3 beta-selection checkpoint. These results provide an important advance for further understanding of thymus biology, especially in view of the contribution of heterogenous thymic macrophage subpopulations.

However, Zhou et al. previously reported essentially similar heterogeneity in thymic macrophages. It was demonstrated that TIMD4+ macrophages and CX3CR1+ macrophages have distinct origins and are different in developmental characteristics (27). The authors should better clarify what was previously demonstrated and what is newly described in this study. Zhou, et al. also demonstrated that TIMD4+ macrophages are localized in the cortex whereas CX3CR1+ macrophages distribute in the medullary region. Whether or not these previous findings are reproduced and supported in the present study is important in view of the new finding that thymic macrophages are important for beta-selection, which is presumed to occur in the thymic cortex. The authors may be able to suggest more strongly that TIMD4+ macrophages regulate beta-selection in the thymic cortex through phagocytic efferocytosis. (Indeed, the Figure 1 legend states that frozen thymic sections were used for immunofluorescent staining to identify the localization of thymic macrophages, without showing the results.)

Author response:

The following is the authors’ response to the original reviews.

We thank the editors and reviewers for their thoughtful evaluation and helpful recommendations. These comments helped us distinguish more clearly between two complementary advances in the study. First, the phenotypic, transcriptomic, and ontogenetic analyses refine the organization of the thymic macrophage compartment and identify CCR2 dependence of the TIMD4- VCAM1+ population. Second, the MaFIA fetal thymus organ culture (FTOC) experiments reveal that an intact Csf1r-expressing myeloid compartment is required for efficient progression across the DN3-to-DN4 checkpoint. The revised manuscript now presents both advances directly while matching the cellular specificity of each conclusion to the experimental system that supports it.

The central functional result is supported by a coordinated set of observations: depletion of Csf1r expressing cells reduces CD4+CD8+ double-positive (DP) ab-lineage thymocyte production, preserves γδ T cell output, causes reciprocal accumulation of DN3 and loss of DN4 cells, and reduces CD27 expression within the DN compartment. We now emphasize that convergence, which localizes the phenotype to the b-selection transition. At the same time, because the MaFIA system targets Csf1r-expressing myeloid cells rather than a single macrophage subset, the manuscript assigns the demonstrated requirement to the thymic myeloid compartment, this now reflected on the revised title as well. This framing preserves the biological importance of the result without attributing more cellular specificity than the experiment provides.

The major revisions include:

- A revised Title and an Abstract that leads with the principal conclusions: TM specialization, unequal developmental contributions, CCR2 dependence, and a requirement for Csf1r expressing myeloid cells during DN3-to-DN4 progression.

- A clearer account of what Zhou et al. established and how the present work adds VCAM1-based prospective resolution, intravascular-labeling information, SpiC analysis, CCR2 dependence, and functional fetal thymus organ culture (FTOC) data.

- A Results section that explains the convergence of DN3, DN4, CD27, DP cells, and γδ T cell measurements as part of developmental transition.

- More precise interpretation of relative marker expression, pseudotime, SpiC deficiency, Tomato/GFP double-positive events, and thymocyte-associated transcripts in the scRNA-seq results.

- A focused Discussion paragraph that acknowledges the cellular breadth of the MaFIA model while retaining the conclusion that the Csf1r-expressing myeloid niche supports early ab T-lineage development.

- Expanded figure legends that make the gating, reporter, heatmap, MaFIA construct, and developmental interpretations easier to follow.

The revision is based on fuller analysis and clearer presentation of the results and datasets.

eLife Assessment

The macrophage characterisation is interesting, although the evidence for the specific involvement of macrophages in beta-selection is incomplete, as alternative explanations have not been ruled out.

We agree that the depletion experiment resolves a requirement for the Csf1r-expressing thymic myeloid compartment rather than for macrophages alone. We have revised the manuscript with this distinction in mind. Importantly, the developmental conclusion remains strong: DN3 accumulation, DN4 loss, reduced CD27 expression, and diminished CD4+CD8+ double-positive (DP) output all point to impaired progression at the b-selection checkpoint, while preserved γδ T cell output and epithelial-cell numbers argue against nonspecific failure of the entire fetal thymus organ culture (FTOC). The revised text therefore states that Csf1r-expressing myeloid cells support this early ab T cell checkpoint, while discussing the relative contributions of TMs, monocytes, and DCs as the next level of cellular resolution.

The Title, Abstract, final Results section, Discussion, Conclusion, and Figures 8-10 Legends were revised to make the positive compartment-level conclusion explicit and consistent.

Public Reviews:

Reviewer #1 (Public review):

The thymic macrophage depletion experiments are not well controlled; DCs are also depleted, the fetal thymus has little or no medulla, and direct AP20187 toxicity to DN thymocytes in MaFIA mice has not been excluded.

The reviewer identifies the key issue of cellular attribution. We have recast the experiment according to what the MaFIA system directly tests: the functional contribution of Csf1r-expressing myeloid cells within an intact FTOC. This interpretation is supported by efficient loss of both TM populations, preservation of EpCAM+ epithelial-cell numbers, and absence of TM depletion in AP20187treated non-transgenic C57BL/6 FTOCs. We no longer use adult cortical-versus-medullary anatomy to infer that macrophages must be the sole responsible population in the fetal thymus. Instead, we emphasize the experimentally secure result that perturbing the Csf1r-expressing myeloid niche produces a selective and internally consistent defect in ab T cell development at the DN3-to-DN4 transition. The possibility of contributions from DCs, monocytes, or direct transgene expression in a thymocyte fraction is addressed once, in a focused Discussion paragraph, as the rationale for assigning the conclusion at the compartment level.

The MaFIA Results section now leads with the transgene-dependent depletion and the convergent developmental phenotype; the Discussion contains a balanced statement of cellular resolution; the Title and Legends consistently refer to Csf1r-expressing myeloid cells.

Reviewer #2 (Public review):

Zhou et al. previously reported similar TM heterogeneity, localization, and developmental characteristics. The manuscript should distinguish prior findings from new findings and address localization in relation to beta-selection.

We have made this distinction explicit throughout. Zhou et al. established the two-population framework, their cortical versus medullary/cortico-medullary localization, their broad embryonic versus adult hematopoietic origins, and their age-associated remodeling. Building on that foundation, the present study contributes: (i) a prospective TIMD4/VCAM1 gating strategy linked to MafB and Csf1r reporters; (ii) intravascular-labeling evidence that both VCAM1+ populations are predominantly parenchymal; (iii) transcriptomic definition of efferocytic versus antigen-presentation/interferon programs; (iv) evidence that total TM abundance is maintained independently of SpiC; (v) selective CCR2 dependence of TIMD4- VCAM1+ macrophages and thymic monocytes; and (vi) functional evidence that the Csf1r-expressing thymic myeloid compartment supports DN3-to-DN4 progression. We use the anatomical localization established by Zhou et al. as the relevant biological context and reserve our new conclusions for the endpoints measured here. Accordingly, the FTOC phenotype is assigned to the Csf1r-expressing myeloid compartment rather than specifically to cortical TIMD4+ macrophages.

The Introduction now clearly separates established knowledge from the questions addressed here, and the Results and Discussion explicitly identify the study-specific advances. Reference 27 has also been corrected to the final eLife publication.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Trajectory analysis

The Abstract mentions trajectory analysis, but there is no corresponding Results section.

We added a Results paragraph describing the Monocle3 analysis and integrated its interpretation with the fate-mapping and CCR2 experiments. Rooting the trajectory in Ly6c2+ Ccr2+ monocytes produces a transcriptional continuum toward macrophage states, supporting progressive acquisition of macrophage programs from a monocyte-like state. We also explain why pseudotime is complementary to, but not a substitute for, genetic lineage information: it models transcriptional relationships and therefore is not used to infer direct conversion between the two mature TM populations. This rationale allows the analysis to contribute meaningfully without asking it to resolve ontogeny on its own.

The trajectory Methods were clarified and a new Results paragraph was added after the scRNA-seq specialization analysis; the Abstract now summarizes the result at the appropriate level.

(2) Comparison with the Zhou et al. gating strategy

Compare the TIMD4/VCAM1 gating strategy with the CD64/F4/80/TIMD4 definition used by Zhou et al., including the identity of TIMD4-VCAM1- cells.

We now map the two VCAM1+ gates directly onto the established TM framework. TIMD4+ VCAM1+ cells correspond to the TIMD4+ cortical population, whereas TIMD4- VCAM1+ cells show the CX3CR1 enrichment expected of the medullary/cortico-medullary population. VCAM1 therefore adds a useful prospective discriminator within the CD64+ F4/80+ parent gate. The TIMD4- VCAM1- gate is MafB-low/negative, Csf1r-EGFP-high, and enriched for Ly6C, CCR2, and CX3CR1, supporting its designation as monocyte-enriched. We use “monocyte-enriched” because it accurately captures the dominant phenotype without implying that every event in the gate is developmentally identical.

The first Results section and Figures 1-2 Legends now explain this correspondence and terminology.

(3) Difference in IV-CD45 labeling

Why are approximately 40% of Ly6C+CD11b+ cells labeled, but only approximately 10% of TIMD4-VCAM1- cells?

The percentages arise from different denominators. Ly6C+ CD11b+ is a broad myeloid gate that contains both blood-exposed and parenchymal cells. TIMD4- VCAM1- is a narrower population defined within the CD64+ F4/80+ parent gate and therefore samples a different compartment. We now make that gating relationship explicit. The i.v.-labeling experiment consequently supports two positive conclusions: both VCAM1+ macrophage populations are predominantly parenchymal, and the broad Ly6C+ CD11b+ compartment contains a substantially larger blood-exposed component.

The i.v.-labeling Results paragraph and Figure 2 Legend now describe the two gates and their distinct denominators.

(4) SpiC requirement in individual TM populations

Determine proportions and numbers of TM subpopulations in Spic-/- mice.

The biological rationale for this suggestion is strong because Spic is enriched in TIMD4+ VCAM1+ macrophages. Figure 3C, however, quantifies the aggregate CD64+ F4/80+ TM compartment. We therefore revised the conclusion to the level directly supported by that experiment: total TM abundance is maintained in Spic-/- mice despite the expected loss of splenic red pulp macrophages. This is an informative distinction because it shows that the overall thymic macrophage compartment does not share the obligate SpiC dependence of red pulp macrophages, even though a subset-selective quantitative or functional effect remains a question for future work.

The Figure 3 Results paragraph, Discussion, and legend now state preservation of total TM abundance rather than making a population-by-population claim.

(5) Tomato/GFP double-positive cells

Many cells appear to express both GFP and tdTomato. What does this mean?

We expanded the explanation of the mTmG reporter. Flt3-Cre-mediated recombination initiates a switch from membrane Tomato to membrane GFP, but the pre-existing membrane Tomato protein need not disappear instantaneously. Tomato+ GFP+ events are therefore consistent with recent/incomplete reporter transition or persistence of stable Tomato protein after recombination. We do not treat them as a third ontogenetic lineage. The key comparative result is the distribution of reporter histories across populations: TIMD4+ VCAM1+ macrophages retain a large FLT3-independent fraction, whereas TIMD4- VCAM1+ macrophages and monocytes show substantially greater FLT3 history.

The fate-mapping Results paragraph and Figure 6 Legend now define the reporter transition and the interpretation of double-positive events.

(6) Direct AP20187 toxicity in DN thymocytes

Test AP20187 toxicity in MaFIA and control DN thymocytes, for example by active caspase-3 staining.

The proposed caspase-3 analysis is designed to determine whether a DN thymocyte fraction expresses sufficient MaFIA transgene to be directly affected. The existing controls establish two important features of the result: AP20187 does not reduce TM numbers in non-transgenic C57BL/6 FTOCs, demonstrating transgene dependence, and the developmental response is patterned rather than global, with preserved γδ T cell output, DN3 accumulation, DN4 loss, and reduced CD27 expression. These convergent observations support the conclusion that integrity of the Csf1r-expressing myeloid compartment is required for efficient DN3-to-DN4 progression. We now state that conclusion prominently and note the remaining question of thymocyte-intrinsic transgene activity once, in the focused Discussion paragraph.

The MaFIA Results now emphasize the transgene-dependent control and convergent developmental measurements; the Discussion states the remaining cellular-resolution issue in one focused paragraph.

Reviewer #2 (Recommendations for the authors):

(1) Figure 1 color annotation

Clarify whether the color annotation in panel C is applied to panel F.

Yes. Panel F evaluates MafB-mCherry and Csf1r-EGFP reporter expression within the same TIMD4/VCAM1-defined populations shown in panel C. We now state this explicitly so the reporter patterns can be interpreted as independent support for the identity of each prospectively defined gate. We also removed the previous reference to immunofluorescence localization because localization data are not displayed in Figure 1.

Figure 1 Legend revised.

(2) Figure 2 controls, thresholds, and population identities

Provide staining controls and thresholds; reconsider categorical “lack” statements; clarify Ly6C+CD11b+ and CD64+F4/80+ populations and the apparent IV-CD45 discrepancies.

The underlying interpretive point is whether marker expression is categorical or relative. We revised the text to describe Ly6C, CCR2, and CX3CR1 comparatively, which more faithfully represents continuous flow-cytometric measurements. This improves the biological conclusion: both VCAM1+ populations are Ly6C-low/negative and CCR2-low relative to the TIMD4- VCAM1- monocyte-enriched population, while CX3CR1 is enriched in TIMD4- VCAM1+ relative to TIMD4+ VCAM1+ macrophages. We also define the broad Ly6C+ CD11b+ comparison gate, the CD64+ F4/80+ macrophage parent gate, and the narrower TIMD4/VCAM1 subgates. With those denominators made explicit, the iv-CD45 measurements become internally consistent rather than apparently contradictory.

The Figure 2 Results paragraph and legend now use relative marker language and explain the gate hierarchy.

(3) Figure 3 colors, VCAM1, and SpiC populations

Explain panel B colors and the apparent Vcam1 difference; provide individual TM-population data in Spic-/- mice.

Panel B is a row-scaled expression heatmap: yellow and purple indicate relatively higher and lower scaled expression for each gene, respectively. These colors should not be read as absolute expression or compared directly with antibody fluorescence, because transcript abundance and cell-surface protein are regulated at different levels and have different dynamic ranges. The legend now makes this distinction explicit. Vcam1 transcript enrichment is nevertheless consistent with VCAM1 protein being a useful surface discriminator in our gating scheme. For SpiC, we now limit the conclusion to the measurement displayed in panel C, as preservation of total CD64+ F4/80+ TMs, while explaining the biological significance of their divergence from SpiC-dependent splenic red pulp macrophages.

Figure 3 Results, Discussion, and Legend revised.

(4) Figure 4 genes not shown

H-2K, H-2D, and Runx3 are described but not shown in the figure.

We clarified the division of information between Figure 4 and Table 1. Figure 4 displays pathway-level GO enrichment, which supports the higher-order conclusion that TIMD4- VCAM1+ macrophages are enriched for antigen-presentation and interferon-response programs. Individual genes contributing to the subset signatures, including H2-K1, H2-D1, and Runx3, are reported in Table 1. The revised wording no longer implies that those individual genes are plotted in Figure 4.

Figure 4 Results paragraph and Legend revised.

(5) Figure 6 double-positive reporter cells

Explain Tomato+GFP+ cells and their relationship to single-positive cells.

As described in our response to Reviewer 1, we now explain the kinetics of the mTmG reporter switch and interpret double-positive events as reporter-transition/persistence events rather than a separate lineage. This interpretation focuses the analysis on the biologically informative comparison— the different balance of FLT3-independent and FLT3-history labeling among TIMD4+ VCAM1+, TIMD4- VCAM1+, and monocyte-enriched populations.

Fate-mapping Results and Figure 6 Legend revised.

(6) CX3CR1 phenotype in Figures 6 and 7

The text describes CX3CR1 phenotype without showing the data.

We now describe Figures 6 and 7 using the markers actually displayed in those panels: TIMD4 and VCAM1. The relationship to CX3CR1 is established independently in Figure 2 and in the integrated scRNA-seq analysis. This separation makes the evidentiary chain clearer: Figure 2 links the surface-defined populations to CX3CR1 phenotype, Figure 6 compares their FLT3 reporter histories, and Figure 7 tests their CCR2 dependence.

Corresponding Results language and Figures 6-7 Legends revised.

(7) Figure 8 Csf1r-EGFP in thymocytes/DCs and “45.2”

Show Csf1r-EGFP expression in thymocytes and DCs; determine whether their reduction is indirect or direct; clarify 45.2.

The reviewer highlights why the MaFIA result should be interpreted at the Csf1r-expressingcompartment level. The reporter is demonstrably expressed by both TM populations and thymic monocytes, and AP20187 causes transgene-dependent TM depletion in FTOC. DC reduction may reflect direct transgene activity, dependence on the altered myeloid niche, or both; similarly, direct transgene expression was not measured in fetal DN thymocytes. We now make that cellular resolution explicit while emphasizing the developmental conclusion supported by the complete phenotype. We also clarify that CD45.2 denotes the congenic allele of the C57BL/6 control and not a distinct treatment or numerical value.

MaFIA Results, Discussion, and Figure 8 Legend revised.

(8) Figure 10 CD27 profiles

Show CD27 flow-cytometric profiles.

We revised the text and legend to make clear that Figure 10 reports quantified frequencies of CD27+ DN3 and DN4 cells. The CD27 measurement is interpreted in conjunction with, rather than in isolation from, the DN-stage data. Reduced CD27 among DN3 cells, reciprocal DN3 accumulation and DN4 loss, and reduced downstream DP output form a coherent sequence that localizes the developmental impairment to the b-selection-associated transition. We therefore use CD27 as a correlate of successful pre-TCR-associated progression rather than claiming direct biochemical measurement of pre-TCR signaling.

Final Results paragraph, Discussion, and Figure 10 Legend revised.

(9) Supplemental Figure 2 schemes

Improve the schemes and explain dLNGFR and AP20187.

The revised Figure Legend now explains each functional element. dLNGFR is the membrane-targeting low-affinity nerve growth factor receptor segment within the MaFIA fusion construct and is not used here as a lineage marker. AP20187 is a synthetic homodimerizer that binds the engineered FKBP domains, bringing the Fas intracellular domains together and initiating apoptosis in transgene-expressing cells. These definitions make the logic of the depletion system understandable without requiring familiarity with the original MaFIA construct.

Supplemental Figure 2 Legend revised.

Closing statement

The revised manuscript now presents the study in clear, evidence-matched terms. It identifies the advances in TM phenotypic organization, transcriptional specialization, developmental contribution, SpiC independence of total TM abundance, and CCR2 dependence of the TIMD4- VCAM1+ population. It also emphasizes the convergent evidence that Csf1r-expressing myeloid cells support progression through the DN3-to-DN4 checkpoint, while accurately defining the cellular resolution of the MaFIA experiment. We thank the editors and reviewers for helping us sharpen both the significance and the precision of these conclusions.

Version 2

Abstract

Thymic macrophages (TMs) maintain tissue homeostasis by clearing the large numbers of apoptotic cells generated during T cell development, but how TM heterogeneity relates to their developmental origin and role in thymocyte maturation remains incompletely understood. Using complementary flow-cytometric, single-cell transcriptomic, and genetic approaches, we resolved two major TM populations corresponding to TIMD4+ cortical and CX3CR1+ medullary/cortico-medullary macrophages. TIMD4+ VCAM1+ TMs displayed a prominent efferocytosis and apoptotic-cell-clearance program, whereas TIMD4− VCAM1+ TMs were enriched for antigen-presentation and interferon-response pathways. Fate mapping revealed unequal progenitor contributions to these populations, and CCR2 deficiency selectively reduced TIMD4− VCAM1+ TMs and thymic monocytes, supporting ongoing input from circulating precursors. Exploratory pseudotime analysis further identified a transcriptional continuum from Ly6c2+ Ccr2+ monocytes toward macrophage states. Using MaFIA fetal thymic organ cultures, AP20187-mediated depletion of Csf1r-expressing myeloid cells reduced CD4+CD8+ thymocyte differentiation and produced a coordinated accumulation of DN3 cells, loss of DN4 cells, and reduction in CD27 expression. These convergent changes identify the DN3-to-DN4 transition as a developmental stage that requires an intact Csf1r-expressing myeloid compartment, and establish a functional connection between the thymic myeloid niche and early αβ T cell development. Together, our study refines the phenotypic and developmental organization of mouse TMs and reveals a previously under-appreciated requirement for myeloid-cell support during progression through the β-selection checkpoint.

Introduction

The thymus is the primary lymphoid organ dedicated to the generation of a diverse and self-tolerant repertoire of T lymphocytes, a process fundamental to adaptive immunity (1). This intricate T cell developmental program unfolds within a highly organized microenvironment composed of thymic epithelial cells (TECs), endothelial cells, and various hematopoietic populations, which collectively provide the essential signals for thymocyte differentiation, proliferation, and selection (2–4). The developmental journey of a T cell begins when bone marrow-derived thymic seeding progenitors (TSPs) enter the thymus at the cortico-medullary junction (CMJ) and commit to the T cell lineage, initiating a multi-stage differentiation cascade (5–7).

This journey is punctuated by a series of stringent checkpoints that assess the integrity and functionality of the nascent T cell receptor (TCR) repertoire (8,9). The earliest stages occur in the thymic cortex, where progenitors, lacking expression of CD4 and CD8, termed double-negative (DN) thymocytes (6), start to differentiate, which is characterized by the expression of CD44 and CD25 (3,4). Progression from the DN1 (CD44+CD25−) to the DN2 (CD44+CD25+) and DN3 (CD44−CD25+) stages is marked by the initiation of V(D)J recombination at the Tcrb, Tcrg, and Tcrd gene loci (1). The DN3 stage represents the first critical checkpoint in αβ T cell development, known as β-selection (12,13). Here, thymocytes that have successfully rearranged a functional TCRβ chain pair it with the invariant pre-TCRα chain (pTα) to form the pre-TCR complex. Successful signaling through this complex provides vital survival, proliferation, and differentiation cues, allowing the cell to progress to the DN4 stage and subsequently to the CD4+CD8+ double-positive (DP) stage (14,15). Thymocytes failing to generate a functional pre-TCR are eliminated via apoptosis, ensuring that only cells with a viable TCRβ chain continue their developmental progression (14).

Following β-selection, DP thymocytes undergo further TCRα chain rearrangement and are subjected to two additional selection events. Positive selection ensures that the fully assembled αβ TCR can recognize self-peptides presented by major histocompatibility complex (MHC) molecules on cortical (c)TECs, a process essential for TCR MHC-restriction (16–18). Subsequently, negative selection in the medulla eliminates thymocytes whose TCRs bind with high affinity to self-peptides presented by medullary (m)TECs and dendritic cells (DCs), thereby establishing central tolerance (19,20). These selection processes are notably stringent, resulting in the apoptotic death of over 95% of all developing thymocytes (21).

Integral to this journey and processes are thymic macrophages (TMs), which have long been recognized for their critical homeostatic function. Residing in both the cortex and medulla, TMs are exceptionally efficient phagocytes, responsible for the swift and silent clearance of the millions of apoptotic thymocytes generated daily (22–25). This process, known as efferocytosis, is vital for preventing the release of potentially immunogenic and inflammatory intracellular contents and maintaining the integrity of the thymic architecture (26). However, this view of TMs as mere “housekeepers” is rapidly evolving. Recent advances, particularly single-cell RNA sequencing (scRNA-seq), have revealed significant heterogeneity within the TM compartment, suggesting functional specialization (27). Two principal subsets have been identified in the adult mouse thymus: cortical TIMD4+ macrophages, which have potent efferocytosis function, and CX3CR1+ macrophages, which are enriched at the CMJ and express genes associated with antigen presentation, suggesting at a potential role in negative selection (27).

Recent work by Zhou et al. established that adult mouse TMs comprise embryonically derived TIMD4+ macrophages localized predominantly in the cortex and adult hematopoietic stem cell-derived CX3CR1+ macrophages enriched in the medulla and cortico-medullary junction (27). These findings provided an important framework for TM heterogeneity and ontogeny. However, the surface-marker relationships that permit prospective resolution of these populations, the requirements that maintain them, and their contribution to early thymocyte development remain incompletely defined. In particular, β-selection is commonly viewed as a thymocyte-intrinsic checkpoint, even though DN3-to-DN4 progression occurs within a multicellular thymic microenvironment. This raises the possibility that resident myeloid cells provide supportive functions that help developing thymocytes successfully navigate this transition.

Here, we combined high-parameter flow cytometry, single-cell RNA sequencing, genetic fate mapping, CCR2 and SpiC deficiency, and fetal thymic organ culture (FTOC) to define the phenotypic organization, developmental contributions, and functional significance of the thymic myeloid compartment. Our results identify VCAM1 as a useful component of a prospective gating strategy for the two major TM populations, demonstrate CCR2 dependence of the TIMD4-VCAM1+ population, and reveal that depletion of Csf1r-expressing myeloid cells compromises thymocyte progression across the DN3-to-DN4 checkpoint.

Materials and methods

Animals

Macrophage Fas-Induced Apoptosis (MaFIA), MafB-mCherry-Cre, Spic−/−, and C57BL/6J mice were purchased from the Jackson Laboratory. The Flt3-Cre × Rosa-mTmG mice and Flt3-Cre × Rosa-mTmG mice in the Ccr2−/- background were kindly provided by Dr. Slava Epelman (University Health Network, University of Toronto) (28). All animals were bred and maintained in specific pathogen-free conditions at the Preclinical Research Centre of the Sunnybrook Research Institute (SRI). All animal procedures were approved by the Sunnybrook Health Sciences Centre Animal Care Committee (Toronto, Ontario, Canada).

Timed pregnancies and embryonic thymus analysis

Timed pregnancies were set up by housing male mice (8–24 weeks of age) in separate cages and mating them with female mice (8–24 weeks of age) overnight. The following morning, females were separated from the males, and successful matings were identified by the presence of a vaginal plug. The gestational age of the embryos was designated as embryonic day 0.5 (E0.5) on the day the plug was observed. Thymuses were harvested from embryos at E15 and E17, as well as from postnatal mice at 1 day, 1 week, and 3 weeks of age, in addition to adult (6-8 weeks) mice. Following collection, thymuses were subjected to enzymatic digestion to dissociate the tissue into single-cell suspensions. The resulting cells were stained for flow cytometric analysis.

Intravenous CD45 staining

A CD45-Phycoerythrin (PE) antibody (30-F11; BioLegend) was diluted in phosphate buffered saline (PBS) to a final concentration of 1.5 µg/150 µL. The antibody solution was administered via tail vein injection using a 1 mL insulin syringe. Three minutes after the injection, mice were euthanized, and thymuses were harvested. The collected tissue was then processed and stained for flow cytometric analysis.

Cell isolation from thymus

Thymocytes were harvested and subjected to enzymatic digestion using the Spleen Dissociation Kit (Miltenyi Biotec) and the gentleMACS™ Dissociator (Miltenyi Biotec) according to the manufacturer’s instructions. The resulting cell suspensions were filtered through a 70 µm cell strainer to remove debris, resuspended in PBS, and kept on ice until further processing.

Flow cytometry

Single-cell suspensions (1 × 107 cells) were prepared from the thymus and stained in a 96-well conical bottom plate (ThermoFisher Scientific). The staining cocktail included the fixable viability dye eFluor 450 (eBioscience), diluted 1:800 in PBS, anti-CD16/32 to block Fc receptors, and the desired fluorochrome-conjugated antibodies targeting macrophages and T cells (see Supplemental Tables 1 and 2 for details). Cells were initially stained with eFluor 450 in PBS for 15 minutes on ice. After washing with fluorescence-activated cell sorting (FACS) buffer, anti-CD64, anti-CCR2, and anti-VCAM1 antibodies were added to the wells, and cells were stained for an additional 15 minutes on ice. To block non-specific binding, Fc-block antibody was added to the cells 15 minutes prior to the addition of the remaining antibody cocktail, which was incubated with the cells for 30 minutes on ice.

Following staining, cells were washed and resuspended in FACS buffer for flow cytometric acquisition using the FACSymphony™ A5 Cell Analyzer (BD Biosciences). Data were analyzed with FlowJo software (TreeStar). The complete list of antibodies and their corresponding clones is provided in Supplemental Tables 1 and 2.

Cell sorting

Thymuses were harvested from 6-week-old C57BL/6 mice and subjected to enzymatic digestion using the Spleen Dissociation Kit (Miltenyi Biotec) and the gentleMACS™ Dissociator (Miltenyi Biotec) to generate single-cell suspensions. After cell isolation, cells were counted and stained with allophycocyanin (APC)-conjugated anti-mouse CD64 antibody in FACS buffer for 20 minutes on ice (see Supplemental Table 1 for antibody details). Following a wash with FACS buffer, cells were stained with anti-APC MicroBeads (Miltenyi Biotec) for 20 minutes on ice. The cells were then washed again with FACS buffer and loaded onto LS columns (Miltenyi Biotec) for positive selection of CD64+ cells. After selection, the cells were washed and further stained with antibodies against F4/80, TIMD4, Ly6C, and CD45 for 30 minutes on ice (see Supplemental Table 1 for antibody details). The stained cells were sorted using the FACSAria™ Fusion Flow Cytometer (BD Biosciences). CD64+ sorted cells were collected in 1.5 mL Eppendorf tubes containing 300 µL of FBS, counted, and resuspended in PBS supplemented with 0.04% BSA. Cells were then encapsulated, and library preparation was performed using the Chromium Single Cell 3’ Reagent Kit (10X Genomics). Gene expression libraries were sequenced on a NovaSeq 6000 platform (Illumina) at Azenta Life Sciences. In total, 3,214 and 2,815 cells were sequenced from two independent experiments.

Fetal thymus organ culture (FTOC) and TM depletion in MaFIA mice

Timed pregnancies were set up for MaFIA mice as described above. Fetal thymuses were harvested at E15.5 for FTOC. Briefly, culture wells were set up in a 12-well-plate 24 hour before FTOC. In each well, Whatman® Nuclepore™ Track-Etched membrane (WHA110409; Sigma) was placed on top of SURGIFORM® absorbable gelatin sponge (1974; Ethicon) in 1.5 mL of complete Dulbecco’s Modified Eagle Medium (DMEM) (supplemented with 10% Fetal Bovine Serum, 2 mM L-glutamine, 100 U/mL penicillin and 100 μg/mL streptomycin) (29,30). Three fetal thymuses were placed on top of the membrane for each well. B/B homodimerizer stock was diluted to 2.5 μM in complete DMEM and added to the thymus lobes at the time of culture setup and again on the following day. On day 2, the membrane with the thymus lobes was transferred to a fresh culture plate containing new gelatin sponge and complete DMEM. Cultures were maintained in 37 °C incubator until day 6, at which point the thymic lobes were processed for flow cytometry analysis as described above.

Transcriptomic analysis

CD64+ cells were isolated from the thymus of C57BL/6J adult mice and processed for scRNA-seq as described below. Data from each scRNA-seq experiment underwent pre-processing as detailed below. Quality control filters were applied to remove cells with more than 12.5% mitochondrial gene expression, fewer than 500 reads, or more than 6,000 reads, ensuring the retention of high-quality cells for analysis. Macrophages and monocytes were identified using a module scoring approach, and their respective cell barcodes were recorded. These barcodes were used for subsetting before merging the datasets from the two scRNA-seq experiments with published TM data from Dzhagalov’s group. The merging was performed using Harmony to align the datasets and correct for batch effects. Following integration, cells were clustered, and module scores were assigned to all identified clusters. Non-relevant cell types, including B cells, T cells, and DCs, were excluded from subsequent analyses to focus on macrophages and monocytes.

Analysis of scRNA-Seq and data integration

All analyses were performed using R software version 4.4.1. Cells expressing fewer than 500 genes or more than 6,000 genes were excluded from the dataset (GEO: GSE326594). Additionally, cells with greater than 12.5% mitochondrial gene content were removed to ensure data quality. The dimensionality of the dataset was assessed using an Elbow plot to determine the appropriate number of principal components. Gene ontology (GO) enrichment analysis of differentially expressed genes (DEGs) was conducted to identify overrepresented biological processes using the FindMarkers function from the Seurat package (31). The data were then integrated with previously published scRNA-seq datasets on TMs (GEO: GSE185460), as well as datasets from heart, liver, and lung macrophages (GEO: GSE188647), from the laboratories of Drs. Dzhagalov and Epelman, respectively. Batch correction across datasets was performed using the Harmony package (27,32,33). Following integration, the merged dataset was normalized using the SCTransform method to account for technical variation. For data visualization, Uniform Manifold Approximation and Projection (UMAP) was applied to reduce dimensionality and enable visualization of cellular clustering patterns.

Trajectory analysis using Monocle3

Trajectory analysis was performed using Monocle3 to examine transcriptional relationships among monocytes and TMs (34). Following SCTransform normalization and integration in Seurat, the expression matrix, cell metadata, and gene metadata were used to construct a Monocle3 CellDataSet. Principal component analysis and UMAP were used for dimensionality reduction, followed by re-clustering and graph learning. Ly6c2+ Ccr2+ monocytes were assigned as the root population based on their established precursor phenotype. Pseudotime therefore represents a model of transcriptional progression from a monocyte-like state and is interpreted together with, rather than independently of, the genetic fate-mapping and CCR2-dependence data.

Statistical analysis

Data are presented as the mean ± standard error of the mean (SEM). Univariate comparisons between two groups were performed using the Mann-Whitney U test. For analyses involving more than two groups, the normality of the data was assessed using the D’Agostino-Pearson test and/or inspection of Q-Q plots, while homoscedasticity (equal variance) was evaluated using the Brown-Forsythe test or by examining residual plots. Non-parametric comparisons across multiple groups were conducted using the Kruskal-Wallis test, followed by post-hoc pairwise comparisons of the mean ranks of each group with the mean rank of the control group. Multiple comparisons were adjusted using Dunn’s test to control for type I error. For pairwise comparisons involving more than two groups, the Holm-Šídák method was applied to correct for multiple comparisons, reducing the likelihood of type I error. All statistical analyses were performed using GraphPad Prism version 10.

Results

Identification of Two Distinct Tissue-Resident Macrophage Subsets in the Adult Thymus

To investigate the heterogeneity of phagocytic cells within the thymus, we performed multi-parameter flow cytometry on hematopoietic cells (CD45+) from the thymus of adult mice. We first identified the total TM population based on the co-expression of the canonical markers CD64 and F4/80 (Figure 1A-B). The majority of these CD64+ F4/80+ cells also expressed the phagocytic receptor MerTK, consistent with their identity as macrophages (35).

Flow-cytometric characterization of thymic macrophage populations.

Thymuses from adult C57BL/6 mice were enzymatically digested and analyzed by flow cytometry. A) Gating strategy for CD64+ F4/80+ TMs. B) Representative MerTK histogram for TMs (blue) and CD64-F4/80-lymphocytes (red). C) TIMD4/VCAM1 definition of TIMD4+ VCAM1+, TIMD4-VCAM1+, and TIMD4-VCAM1-populations. D) Frequencies of the three TM populations. E) MerTK+ frequencies. F) MafB-mCherry and Csf1r-EGFP reporter expression evaluated within the same TIMD4/VCAM1 gates defined in panel C. Data are mean ± SEM (n = 7 per group from two independent experiments). Kruskal-Wallis test (*p ≤ 0.05).

Analysis of TIMD4 and VCAM1 expression resolved three populations within the CD64+ F4/80+ gate: a major TIMD4+ VCAM1+ population, a smaller TIMD4-VCAM1+ population, and a minor TIMD4-VCAM1-population (Figure 1C-E). The two VCAM1+ gates correspond phenotypically to the TIMD4+ cortical and TIMD4-/CX3CR1+ medullary/cortico-medullary macrophage populations described by Zhou et al., while VCAM1 provides an additional prospective discriminator (27). Both VCAM1+ populations expressed the macrophage-associated MafB-mCherry reporter. By contrast, the TIMD4-VCAM1-population was MafB-low/negative and displayed high Csf1r-EGFP, consistent with a monocyte-enriched phenotype (Figure 1F). These results establish a flow-cytometric framework that links our prospectively defined populations to the major TM states identified transcriptionally and anatomically in prior work (27).

Short-term intravenous (i.v.) labeling with anti-CD45 demonstrated that most TIMD4+ VCAM1+ and TIMD4-VCAM1+ cells were shielded from the circulation, supporting their predominantly parenchymal localization (Figure 2A-C). The broad Ly6C+ CD11b+ comparison gate contained both iv-labeled and unlabeled cells, whereas the more stringently defined TIMD4-VCAM1-gate within CD64+ F4/80+ cells contained a smaller i.v.-labeled fraction. The two VCAM1+ populations were Ly6C-low/negative and expressed substantially less CCR2 than TIMD4-VCAM1-cells. CX3CR1 expression was enriched in the TIMD4-VCAM1+ population relative to TIMD4+ VCAM1+ macrophages (Figure 2D-E), further aligning these populations with the TIMD4+ and CX3CR1+ TM states described previously (27). In contrast, the TIMD4-VCAM1-population was enriched for Ly6C, CCR2, and CX3CR1, supporting its designation as a monocyte-enriched thymic population.

Intravascular labeling and monocyte-associated marker expression distinguish thymic myeloid populations.

A) Short-term intravenous (i.v.) anti-CD45 labeling identifies blood-exposed cells. B) Frequencies of i.v. CD45-labeled cells in the broad Ly6C+ CD11b+ gate and the CD64+ F4/80+ macrophage gate. The Ly6C+ CD11b+ gate contains heterogeneous blood-exposed and parenchymal cells and is not equivalent to the TIMD4-VCAM1-gate. C) i.v. CD45 labeling within TIMD4/VCAM1-defined populations. D-E) Relative Ly6C, CCR2, and CX3CR1 signal across the three populations; descriptions are comparative rather than categorical. Data are mean ± SEM (n = 5-7 per group from two independent experiments). Mann-Whitney test for two-group comparisons and Kruskal-Wallis test for three-group comparisons (*p ≤ 0.05; **p ≤ 0.01).

Together, these data identify two predominantly parenchymal VCAM1+ TM populations and distinguish them from a TIMD4-VCAM1-monocyte-enriched population. This gating framework connects surface phenotype, anatomical accessibility, and reporter expression, enabling prospective analysis of the major TM populations.

Transcriptomic Profiling by scRNA-seq Confirms Two Functionally Distinct Macrophage Subsets

To gain deeper insight into the transcriptional identities of these macrophage populations, we performed scRNA-seq on sorted thymic CD45+ CD64+ cells. Unbiased clustering of the resulting 3,153 cells revealed five distinct myeloid clusters, including monocytes, proliferating macrophages, and two major quiescent macrophage populations that corresponded to our flow-cytometric gates (Figure 3A).

Single-cell transcriptomic profiles of thymic macrophages.

CD64+ cells from six-week-old C57BL/6 thymuses were analyzed by 10X Genomics 3’ scRNA-seq in two independent experiments, each pooling six mice, and integrated with the Zhou et al. dataset using Harmony. A) UMAP showing five annotated myeloid clusters; panel colors denote cluster identity. B) Heatmap of row-scaled differentially expressed genes; yellow and purple represent higher and lower relative scaled expression, respectively, and are not equivalent to flow-cytometric fluorescence intensity. C) Total CD64+ F4/80+ macrophage numbers in thymus and spleen of Spic-/- and control mice; TM subpopulations were not separately quantified in this panel. Data are mean ± SEM (n = 4-5 per group from two independent experiments), Mann-Whitney test (*p ≤ 0.05).

One major macrophage cluster expressed Timd4, Vcam1, Mertk, and Maf, corresponding to the TIMD4+ VCAM1+ population (Figure 3B). A second cluster expressed Vcam1 and Cx3cr1, aligning with the TIMD4-VCAM1+ population (Table 1). The TIMD4+ VCAM1+ cluster also expressed Spic, the lineage-defining transcription factor required for splenic red pulp macrophages (RPMs) (36). We therefore examined the TM compartment in Spic-/- mice. As expected, splenic RPMs were markedly reduced, whereas total CD64+ F4/80+ TM abundance was preserved (Figure 3C). Thus, despite sharing Spic expression with RPMs, maintenance of the overall TM compartment follows a distinct developmental program that does not require SpiC.

Gene Ontology analysis revealed pronounced functional specialization between the two TM populations. TIMD4-VCAM1+ CX3CR1+ TMs were enriched for antigen processing and presentation via MHC class I and for interferon-response pathways, consistent with a potential role in immune surveillance and T cell education (Figure 4A). Conversely, TIMD4+ VCAM1+ TMs were strongly enriched for phagocytosis and endocytosis pathways, supporting specialization for efferocytic clearance of apoptotic thymocytes (Figure 4B). Individual genes contributing to these subset signatures, including H2-K1, H2-D1, and Runx3, are reported in Table 1.

Gene Ontology pathway analysis of TM populations.

A) Top enriched pathways in TIMD4-VCAM1+ CX3CR1+ TMs. B) Top enriched pathways in TIMD4+ VCAM1+ TMs. The panels display pathway-level enrichment; individual genes mentioned in the text are reported in Table 1 and are not plotted separately in this Figure.

A subcluster within the TIMD4+ VCAM1+ population co-expressed macrophage genes together with thymocyte-associated transcripts, including Cd3g and Trac (Figure 3B). In the context of the strong phagocytosis and endocytosis program in this population, the presence of thymocyte-derived transcripts is consistent with ongoing uptake of thymocyte material and provides additional support for efferocytic specialization. As with other phagocyte single-cell datasets, contributions from ambient RNA or rare multiplets cannot be fully distinguished from engulfed RNA, and we therefore interpret this signature together with the broader pathway evidence.

Exploratory Monocle3 analysis, rooted in Ly6c2+ Ccr2+ monocytes, placed monocytes and macrophage states along a connected transcriptional manifold. The ordering was consistent with progressive acquisition of macrophage-associated programs from a monocyte-like state and provided a transcriptional complement to the fate-mapping and CCR2-dependence experiments. As pseudotime models transcriptional relationships rather than prospective cell fate, it only allows us to predict a potential conversion between the two mature TM populations.

The Thymic Macrophage Compartment Undergoes Dynamic Remodeling During Development

To understand how the TM compartment evolves in parallel with the thymus itself, we analyzed the number and composition of TM subsets during ontogeny, from embryonic development through adulthood. Consistent with known thymic biology, the absolute number of total thymocytes and TMs peaked in young, 3-week-old mice, before declining with age, indicating that the size of the TM pool is coupled to the overall cellularity of the organ (Supplemental Figure 1A-B).

TM composition was extensively remodeled across development. At embryonic day 15 (E15), the thymic myeloid compartment was dominated by TIMD4-VCAM1-cells expressing CD11b and CCR2, consistent with recently recruited or immature myeloid precursors (Figure 5A).

Developmental progression of TM subsets.

Thymuses were harvested from mice at various developmental time points and enzymatically digested to prepare single-cell suspensions. Cells were stained with antibodies for flow cytometry analysis. A) Expression of TM markers across different age groups. B) Comparison of the numbers and percentages of TM in mice at different ages. Data are presented as mean ± SEM (n = 4-12 per group from 2-4 independent experiments).

Following birth, TIMD4+ VCAM1+ macrophages expanded rapidly and became the dominant TM population during the period of maximal thymic growth (Figure 5B). With the onset of age-associated thymic involution, the proportion of TIMD4+ VCAM1+ macrophages declined, whereas TIMD4-VCAM1+ macrophages progressively increased. These data reproduce the age-associated shift reported by Zhou et al. and show that our VCAM1-based gating strategy captures the developmental remodeling of the two major TM populations (27).

Thus, the thymic macrophage compartment is developmentally dynamic: immature myeloid cells predominate in the fetal organ, TIMD4+ VCAM1+ macrophages expand with postnatal thymic growth, and TIMD4-VCAM1+ macrophages become increasingly prominent during involution. The changing abundance of these populations establishes the developmental context for their distinct ontogenetic contributions.

Thymic Macrophage Subsets Arise from Distinct Developmental Origins

Flt3-Cre × Rosa-mTmG fate mapping revealed unequal developmental contributions to the two TM populations. In this model, cells with a history of FLT3 expression switch from membrane Tomato to membrane GFP, whereas FLT3-independent cells remain Tomato+ (Figure 6A) (28). TIMD4+ VCAM1+ macrophages contained a large Tomato+ fraction, demonstrating a substantial contribution from FLT3-independent embryonic/fetal precursors. In contrast, TIMD4-VCAM1+ macrophages and thymic monocytes were predominantly GFP+, consistent with derivation from FLT3-expressing hematopoietic progenitors (Figure 6B-C). Tomato+ GFP+ events likely represent ongoing reporter transition or persistence of stable membrane Tomato after recombination and were included within the overall assessment of reporter history rather than treated as a separate lineage.

Flt3-Cre reporter history in thymic myeloid populations.

A) In Flt3-Cre × Rosa-mTmG mice, Cre-mediated recombination switches membrane Tomato expression to membrane GFP in cells with Flt3 expression history. B) Representative Tomato/GFP profiles within TIMD4+ VCAM1+, TIMD4-VCAM1+, and TIMD4-VCAM1-gates. Tomato+ GFP+ events are interpreted as reporter-transition/persistence events and not as a distinct lineage. C) Reporter frequencies among the three populations. CX3CR1 phenotype is defined independently in Figure 2 and is not displayed in this Figure. Data are mean ± SEM (n = 7 per group from two independent experiments). Mann-Whitney tests with Holm-Šídák correction (***p ≤ 0.001).

Because TIMD4-VCAM1+ macrophages showed greater FLT3-lineage history than their TIMD4+ VCAM1+ counterparts, we asked whether maintenance of this population depends on CCR2-mediated precursor recruitment. We therefore analyzed thymic myeloid populations in Ccr2 knockout (Ccr2-/-) mice. Overall thymocyte cellularity and the abundance of TIMD4+ VCAM1+ macrophages were preserved, whereas both TIMD4-VCAM1+ macrophages and the TIMD4-VCAM1-monocyte-enriched population were significantly reduced (Figure 7A-D). This selective effect links CCR2-dependent recruitment to maintenance of the TIMD4-VCAM1+ arm of the thymic macrophage network.

CCR2 dependence of TIMD4/VCAM1-defined thymic myeloid populations.

A) Representative TIMD4/VCAM1 profiles from Ccr2-/- and control mice. B-D) Frequencies and absolute numbers of TIMD4+ VCAM1+, TIMD4-VCAM1+, and TIMD4-VCAM1-populations. CX3CR1 phenotype is defined independently in Figure 2 and is not displayed in this Figure. Data are mean ± SEM (n = 6-7 per group from two independent experiments). Two-tailed Mann-Whitney test (*p ≤ 0.05; **p ≤ 0.01; ***p ≤ 0.001).

Together, the fate-mapping and Ccr2-/- data support a dual-contribution model for the TM network. TIMD4+ VCAM1+ macrophages retain a prominent FLT3-independent embryonic/fetal contribution, whereas TIMD4-VCAM1+ macrophages show greater FLT3 history and depend on CCR2-mediated recruitment. The selective reduction of TIMD4-VCAM1+ macrophages and thymic monocytes in Ccr2-/- mice provides functional evidence that circulating precursors contribute to maintenance of this arm of the TM compartment.

Csf1r-Expressing Myeloid Cells Support Thymocyte Progression Across the DN3-to-DN4 Checkpoint

To determine whether Csf1r-expressing myeloid cells contribute to early T cell development, we used the MaFIA (Csf1r-EGFP-NGFR/FKBP1A/TNFRSF6) model, in which AP20187 dimerizes an FKBP-Fas suicide protein and induces apoptosis in transgene-expressing cells (Supplemental Figure 2A-B) (37). Both TM populations and thymic monocytes expressed the Csf1r-EGFP reporter (Supplemental Figure 2C-E), making this system well suited to perturb the thymic myeloid compartment. We applied this experimental approach to E15.5 fetal thymic organ culture (FTOC), which preserves the three-dimensional thymic microenvironment while avoiding systemic consequences of myeloid-cell depletion (38), and treated FTOCs from MaFIA mice with AP20187 at culture initiation (Figure 8A).

Efficient depletion of Csf1r-expressing myeloid cells in MaFIA fetal thymic organ culture.

E15.5 MaFIA thymus lobes were treated with AP20187 at culture initiation and the following day, transferred to fresh FTOC media on day 2, and analyzed on day 6. A) Experimental workflow. B) EpCAM+ CD45-thymic epithelial-cell frequencies and numbers. C) Total thymocyte numbers. D) CD45+ CD11c+ MHC-IIhigh DC numbers. E-F) TM frequencies and absolute numbers in untreated and AP20187-treated MaFIA FTOCs and in AP20187-treated C57BL/6J (CD45.2) control FTOCs. CD45.2 denotes the congenic allele of the C57BL/6J control. G-H) Numbers of TIMD4+ VCAM1+ and TIMD4-VCAM1+ TMs. The preservation of epithelial-cell and control-FTOC TM numbers, together with depletion of both MaFIA TM populations, demonstrates transgene-dependent perturbation of the Csf1r-expressing thymic myeloid compartment. Data are mean +/− SEM (n = 8-11 per group from three independent experiments). Mann-Whitney or Kruskal-Wallis tests as indicated (*p <= 0.05; **p <= 0.01; ***p <= 0.001).

AP20187 treatment preserved the EpCAM+ CD45-thymic epithelial-cell compartment (Figure 8B) while markedly reducing total TMs and both TIMD4/VCAM1-defined TM populations (Figure 8E-H). Importantly, TM abundance was unchanged in AP20187-treated C57BL/6 control FTOCs lacking the MaFIA transgene (Figure 8F), demonstrating that depletion required the transgenic suicide system rather than exposure to the dimerizer alone. Total thymocyte and CD11c+ MHC-IIhigh DC numbers were also reduced (Figure 8C-D). These results establish efficient, transgene-dependent perturbation of the Csf1r-expressing thymic myeloid compartment within an epithelial environment that remained numerically intact.

Depletion of Csf1r-expressing cells impaired αβ T-lineage cell development, leading to a marked reduction in CD4+ CD8+ double-positive (DP) thymocytes, whereas γδ T cell output was preserved (Figure 9A-D). To localize the αβ lineage developmental defect, we examined upstream DN populations (Figure 10A). AP20187-treated MaFIA FTOCs showed a coordinated accumulation of DN3 (CD44-CD25+) cells and reduction of DN4 (CD44-CD25-) cells (Figure 10B). The reciprocal change in sequential developmental populations identifies the DN3-to-DN4 transition as the principal stage affected by the loss of Csf1r-expressing cells.

Csf1r-expressing myeloid cells support alpha-beta T cell development in FTOC.

A) Representative CD4/CD8 profiles. B) Frequencies and absolute numbers of thymocyte populations in untreated and AP20187-treated MaFIA FTOCs. C) CD3+ CD4SP-to-CD3+ CD8SP ratio. D) Frequencies and numbers of γδ T cells. Depletion reduced αβ-lineage DP thymocyte production while preserving γδ T cell output. Data are mean +/− SEM (n = 8-11 per group from three independent experiments; γδ T cells, n = 6-7 from two experiments). Mann-Whitney tests with Holm-Sidak correction (*p <= 0.05; **p <= 0.01; ***p <= 0.001).

Csf1r-expressing myeloid cells support progression across the DN3-to-DN4 checkpoint.

A) Representative CD44/CD25 profiles of DN populations. B) Frequencies and absolute numbers of DN subsets in untreated and AP20187-treated MaFIA FTOCs. C-D) Quantified frequencies of CD27+ DN3 and DN4 cells. The coordinated accumulation of DN3 cells, reduction of DN4 cells, and lower CD27 expression identify impaired progression at the β-selection transition. Data are mean +/− SEM (DN subsets, n = 8-11 from three independent experiments; CD27, n = 6-8 from two experiments). Mann-Whitney tests with Holm-Sidak correction (*p <= 0.05; **p <= 0.01; ***p <= 0.001).

We next examined CD27 expression, which is upregulated as DN3 thymocytes successfully receive pre-TCR-associated signals and progress through β-selection. The frequency of CD27+ DN3 cells was significantly reduced in AP20187-treated MaFIA FTOCs (Figure 10C-D). Considered together, DN3 accumulation, DN4 loss, reduced CD27 expression, and diminished DP cell output, we have strong evidence that Csf1r-expressing myeloid cells support thymocyte differentiation at the β-selection checkpoint. Thus, the thymic myeloid compartment contributes not only to tissue clearance and homeostasis but also to the microenvironmental conditions required for efficient early αβ T cell development.

Discussion

This study defines the mouse TM compartment as a developmentally dynamic network of phenotypically and transcriptionally specialized macrophages and connects the thymic myeloid niche to an early checkpoint in αβ T cell development. By integrating prospective surface-marker definition with single-cell transcriptomics, fate mapping, genetic perturbation, and organ culture, we identify two major TM populations with distinct functional programs and developmental contributions, demonstrate CCR2 dependence of the TIMD4-VCAM1+ population, and show that Csf1r-expressing myeloid cells support efficient DN3-to-DN4 progression.

The TIMD4+ VCAM1+ population displayed a coherent efferocytic identity, including Mertk expression, enrichment of phagocytosis and endocytosis pathways, and detection of thymocyte-associated transcripts within a macrophage subcluster. Although these cells expressed Spic, preservation of total TMs in Spic-/- mice distinguishes maintenance of the thymic compartment from the SpiC-dependent program of splenic red pulp macrophages (36). In contrast, TIMD4-VCAM1+ macrophages were enriched for antigen-presentation and interferon-response pathways, suggesting specialization for immune surveillance and interactions with developing thymocytes. The surface-marker correspondence between these transcriptional states and the cortical TIMD4+ and medullary/cortico-medullary CX3CR1+ populations described by Zhou et al. provides a unified phenotypic and functional framework (27).

The ontogeny experiments further reveal how this functional diversity is maintained. TIMD4+ VCAM1+ macrophages retained a substantial FLT3-independent embryonic/fetal contribution, whereas TIMD4-VCAM1+ macrophages showed greater FLT3 history. The selective loss of TIMD4-VCAM1+ macrophages and thymic monocytes in Ccr2-/- mice provides functional support for continuing recruitment from circulating precursors. These results confirm the broad dual-origin framework established by Zhou et al. (27), and extend it by identifying CCR2 as a requirement for maintenance of the TIMD4-VCAM1+ arm of the network under steady-state conditions.

The functional studies reveal that the early αβ T cell program is sensitive to disruption of the thymic myeloid niche. The combination of DN3 accumulation, DN4 reduction, diminished CD27 expression, and reduced DP output localizes this requirement to progression at or near β-selection (41,42). Preservation of γδ T cells and thymic epithelial cell numbers argues against a uniform collapse of fetal thymic development and instead identifies a selective vulnerability of the αβ-lineage developmental pathway. These findings expand the established homeostatic role of thymic myeloid cells by showing that their presence is linked to efficient navigation of the first TCRβ rearrangement-dependent checkpoint.

Several complementary mechanisms could account for this myeloid-cell requirement. TIMD4+ macrophages are positioned and transcriptionally equipped to clear apoptotic thymocytes efficiently; their depletion could increase local cellular debris or alter the anti-inflammatory signals normally generated during efferocytosis, including TGF-β and IL-10 (51–53). Myeloid cells may also provide survival, metabolic, or receptor-ligand signals that cooperate with pre-TCR and Notch pathways necessary for DN3-to-DN4 progression (43–48). The MaFIA mouse model depletes a Csf1r-expressing compartment that includes TMs, monocytes, and DCs, and our experiments did not directly assess Csf1r-transgene expression or AP20187-induced apoptosis in purified DN thymocytes. Consequently, the current data establish the requirement at the level of the Csf1r-expressing myeloid niche rather than assigning it to one macrophage subset or a single molecular mechanism. This evidence-matched cellular resolution sharpens the developmental conclusion and provides a focused basis for future cell-restricted studies.

In conclusion, this study provides an integrated phenotypic, transcriptional, and developmental map of mouse TMs and demonstrates that the TIMD4-VCAM1+ population depends on CCR2-mediated precursor input. Most importantly, the organ culture experiments identify an intact Csf1r-expressing myeloid compartment as a requirement for efficient progression across the DN3-to-DN4 checkpoint. These findings place thymic macrophage heterogeneity within a broader developmental framework and establish the thymic myeloid niche as an active component of early T cell development.